---
title: "Optimal segmentation"
author: "Toby Dylan Hocking"
output: beamer_presentation
---

```{r opts, echo=FALSE}
knitr::opts_chunk$set(
  echo=FALSE, results=FALSE,
  fig.width=10,
  fig.height=6)
```

# Motivation for optimal detection

- Binary segmentation is a greedy algorithm, so sometimes it chooses a
  changepoint which is sub-optimal for a larger model size.
- Is it possible to compute the changepoints and segments which are
  optimal for each model size? Yes, with dynamic programming.
- Is it desirable to compute the optimal changepoints and segments?
  Yes, see examples in next slides.
  
---

# Example 1: stuck with sub-optimal change
  
```{r}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
})
data(neuroblastoma, package="neuroblastoma")
nb.dt <- data.table(neuroblastoma[["profiles"]])
setkey(nb.dt, profile.id, chromosome)

get_segs_loss <- function(pid, chrom, max.segments=6){
  one <- nb.dt[J(pid, chrom)]
  logratio <- one[["logratio"]]
  binseg <- binsegRcpp::binseg_normal(logratio, max.segments)
  opt <- jointseg::Fpsn(logratio, max.segments)
  cum.end <- cumsum(logratio)
  cum.start <- c(0,cum.end)
  end <- t(opt[["t.est"]])
  start <- rbind(1, end[-max.segments,]+1)
  n.data <- end-start+1
  seg.sum <- cum.end[end]-cum.start[start]
  ldt <- function(algorithm, loss){
    data.table(algorithm=a(algorithm), segments=seq_along(loss), loss)
  }
  a <- function(x)factor(x, c("optimal", "binseg"))
  loss.dt <- rbind(
    ldt("optimal", opt$J.est),
    ldt("binseg", binseg$loss))
  segs.dt <- rbind(
    data.table(
      algorithm=a("optimal"),
      segments=as.integer(col(end)),
      start=as.integer(start),
      end=as.integer(end),
      mean=as.numeric(seg.sum/n.data)
    )[!is.na(mean)],
    data.table(algorithm=a("binseg"), coef(binseg)))
  one[, data.i := .I]
  change.dt <- segs.dt[1 < start]
  list(segment=segs.dt, change=change.dt, loss=loss.dt, data=one)
}

models <- get_segs_loss("565", "X")
algo.sizes <- c(optimal=1.25, binseg=0.5)
gg <- ggplot()+
  geom_text(aes(
    1000, Inf,
    color=algorithm,
    hjust=ifelse(algorithm=="binseg", 0, 1),
    label=sprintf("%s loss=%.2f", algorithm, loss)),
    vjust=1,
    data=models[["loss"]])+
  geom_point(aes(
    data.i, logratio),
    shape=1,
    data=models[["data"]])+
  geom_segment(aes(
    start-0.5, mean,
    color=algorithm,
    size=algorithm,
    xend=end+0.5, yend=mean),
    data=models[["segment"]])+
  geom_vline(aes(
    xintercept=start-0.5,
    size=algorithm,
    color=algorithm),
    data=models[["change"]])+
  scale_size_manual(values=algo.sizes)+
  scale_x_continuous(
    "Position/index in data sequence")+
  scale_y_continuous(
    "logratio (approximate DNA copy number)")+
  facet_grid(segments ~ ., labeller=label_both)
gg
```

---

# Zoom to start

```{r}
gg+coord_cartesian(xlim=c(0, 200))
```

---

# Zoom to center

```{r}
gg+coord_cartesian(xlim=c(2100, 2300))
```

---

# Zoom to end

```{r}
gg+coord_cartesian(xlim=c(3700, 3800))
```

---

# Example 2: missing a small change down

```{r}
models <- get_segs_loss("229", "X")
gg <- ggplot()+
  geom_point(aes(
    data.i, logratio),
    shape=1,
    data=models[["data"]])+
  geom_segment(aes(
    start-0.5, mean,
    color=algorithm,
    size=algorithm,
    xend=end+0.5, yend=mean),
    data=models[["segment"]])+
  geom_vline(aes(
    xintercept=start-0.5,
    size=algorithm,
    color=algorithm),
    data=models[["change"]])+
  geom_text(aes(
    1000, Inf,
    color=algorithm,
    hjust=ifelse(algorithm=="binseg", 0, 1),
    label=sprintf("%s loss=%.2f", algorithm, loss)),
    vjust=1,
    data=models[["loss"]])+
  scale_size_manual(values=algo.sizes)+
  scale_x_continuous(
    "Position/index in data sequence")+
  scale_y_continuous(
    "logratio (approximate DNA copy number)")+
  facet_grid(segments ~ ., labeller=label_both)
gg
```

---

# Zoom to start

```{r}
gg+coord_cartesian(xlim=c(0, 200))
```

---

# Zoom to center

```{r}
gg+coord_cartesian(xlim=c(2000, 2500))
```

---

# Zoom to end

```{r}
gg+coord_cartesian(xlim=c(3700, 3800))
```

---

# Example 3: difficult to detect an outlier

```{r}
models <- get_segs_loss("590", "1")
gg <- ggplot()+
  geom_point(aes(
    data.i, logratio),
    shape=1,
    data=models[["data"]])+
  geom_segment(aes(
    start-0.5, mean,
    color=algorithm,
    size=algorithm,
    xend=end+0.5, yend=mean),
    data=models[["segment"]])+
  geom_vline(aes(
    xintercept=start-0.5,
    size=algorithm,
    color=algorithm),
    data=models[["change"]])+
  geom_label(aes(
    2000, Inf,
    color=algorithm,
    hjust=ifelse(algorithm=="binseg", 0, 1),
    label=sprintf("%s loss=%.2f", algorithm, loss)),
    vjust=1,
    data=models[["loss"]])+
  scale_size_manual(values=algo.sizes)+
  scale_x_continuous(
    "Position/index in data sequence")+
  scale_y_continuous(
    "logratio (approximate DNA copy number)")+
  facet_grid(segments ~ ., labeller=label_both)
gg
```

---

# Zoom X axis to outlier

```{r}
gg+coord_cartesian(xlim=c(5000, 5100))
```

---

# Example 4: stuck with sub-optimal change 

```{r}
models <- get_segs_loss("2", "2", 4)
gg <- ggplot()+
  geom_text(aes(
    200, Inf,
    color=algorithm,
    hjust=ifelse(algorithm=="binseg", 0, 1),
    label=sprintf("%s loss=%.2f", algorithm, loss)),
    vjust=1,
    data=models[["loss"]])+
  geom_point(aes(
    data.i, logratio),
    shape=1,
    data=models[["data"]])+
  geom_segment(aes(
    start-0.5, mean,
    color=algorithm,
    size=algorithm,
    xend=end+0.5, yend=mean),
    data=models[["segment"]])+
  geom_vline(aes(
    xintercept=start-0.5,
    size=algorithm,
    color=algorithm),
    data=models[["change"]])+
  scale_size_manual(values=algo.sizes)+
  scale_x_continuous(
    "Position/index in data sequence")+
  scale_y_continuous(
    "logratio (approximate DNA copy number)")+
  facet_grid(segments ~ ., labeller=label_both)
gg
```

---

# Zoom to changepoint

```{r}
gg+coord_cartesian(xlim=c(50, 100))
```

---

# Dynamic programming for optimal changepoint detection

- We have $n$ data $z_1, \dots, z_n$.
- Fix the number of segments $S\in\{1, 2, \dots, n\}$.
- Optimization variables: $S-1$ changepoints
    $t_1 < \cdots < t_{S-1}$ and $S$ segment means
    $u_1,\dots,u_S\in\mathbb R$ ($t_0=0$, $t_S=n$).
- Statistical model: for every segment $s\in\{1,\dots,S\}$,
    $z_i \stackrel{\text{iid}}{\sim} N(u_s, \sigma^2)$ for every
    data point $i\in(t_{s-1},t_s]$ implies square loss function
    $\ell(u_s, z_i)=(u_s-z_i)^2$ to minimize.

```{r plotcos, dev='tikz', fig.height=4}
one.dt <- nb.dt[J("4", "2")]
max.segments <- 4
logratio <- one.dt[["logratio"]]
opt <- jointseg::Fpsn(logratio, max.segments)
cum.end <- cumsum(logratio)
cum.start <- c(0,cum.end)
end <- t(opt[["t.est"]])
start <- rbind(1, end[-max.segments,]+1)
n.data <- end-start+1
seg.sum <- cum.end[end]-cum.start[start]
segs.dt <- data.table(
  algorithm=a("optimal"),
  segments=as.integer(col(end)),
  start=as.integer(start),
  end=as.integer(end),
  mean=as.numeric(seg.sum/n.data)
)[!is.na(mean)]
one.dt[, data.i := .I]

show.segs <- segs.dt[segments==max.segments]
show.changes <- show.segs[1 < start]
show.changes[, change.i := .I]
show.segs[, seg.i := .I]
model.color <- "blue"
text.size <- 5
ggplot()+
  ggtitle(sprintf(
    "Optimal model for $n=%d$ data and $S=%d$ segments",
    nrow(one.dt), max.segments))+
  theme(text=element_text(size=20))+
  geom_point(aes(
    data.i, logratio),
    shape=1,
    data=one.dt)+
  geom_segment(aes(
    start-0.5, mean,
    xend=end+0.5, yend=mean),
    color=model.color,
    size=1,
    data=show.segs)+
  geom_vline(aes(
    xintercept=start-0.5),
    data=show.changes,
    size=1,
    color=model.color)+
  coord_cartesian(xlim=c(-20, max(one.dt[["data.i"]])))+
  scale_x_continuous(
    "Position/index in data sequence")+
  scale_y_continuous(
    "logratio = noisy DNA\ncopy number measurement")+
  geom_text(aes(
    start+0.5, -Inf,
    label=sprintf("$t_%d=%d$", change.i, start-1)),
    color=model.color,
    size=text.size,
    hjust=0,
    vjust=-0.5,
    data=show.changes)+
  geom_text(aes(
    start-1.5, mean,
    label=sprintf("$u_%d=%.4f$", seg.i, mean)),
    color=model.color,
    size=text.size,
    hjust=1,
    data=show.segs)

```

---

# Maximum likelihood inference for $S$ segments and $n$ data

The best loss for $S$ segments and $n$ data is
```{=latex}
\begin{eqnarray*}
  \mathcal L_{S,n} &=& \min_{\substack{
  \mathbf u\in\mathbb R^{S}
\\
   0=t_0<t_1<\cdots<t_{S-1}<t_S=n
  }} 
    \sum_{s=1}^S\  \sum_{i=t_{s-1}+1}^{t_s} \ell( u_s,  z_i) 
\\
&=&\min_{t_{S-1}}\underbrace{
\min_{\substack{
u_1,\dots,u_{S-1}\\
t_1<\cdots<t_{S-2}
}}
  \sum_{s=1}^{S-1}\  \sum_{i=t_{s-1}+1}^{t_s} \ell( u_s,  z_i)
}_{
\mathcal L_{S-1, t_{S-1}}
} +
\underbrace{\min_{u_S} \sum_{i=t_{S-1}+1}^{t_S=n} \ell( u_S,  z_i)}_{
c_{(t_{S-1}, t_S=n]}
}
%      \text{subject to \hskip 0.75cm} &\ \ \alert<3>{u_{s-1} \leq u_s\ \forall s\in\{2,4,\dots\},}  \nonumber\\
%  &\ \ \alert<3>{u_{s-1} \geq u_s\ \forall s\in\{3,5,\dots\}.}  \nonumber 
\end{eqnarray*}
```

- Hard optimization problem because of integer-valued changepoint $t_s$
  variables, naively $O(n^S)$ time.
- Auger and Lawrence (1989): $O(Sn^2)$ time classical dynamic
  programming algorithm (best loss computed recursively).
$$
\mathcal L_{s,t}=\min_{t'<t} \mathcal L_{s-1,t'} + c_{(t', t]}
$$

---

# First step of dynamic programming

```{r}
N.data <- length(logratio)
loss.mat <- matrix(NA, N.data, max.segments)
cum.vec <- cumsum(logratio)
cum.sq.vec <- cumsum(logratio^2)
N.vec <- 1:N.data
loss.mat[,1] <- cum.sq.vec-cum.vec^2/N.vec
loss.dt <- data.table(
  data.i=N.vec,
  mean=cum.vec/N.vec,
  loss=loss.mat[,1])
gg <- ggplot()+
  scale_x_continuous(
    "Position/index in data sequence")+
  scale_y_continuous("")+
  geom_point(aes(
    data.i, logratio),
    shape=1,
    data=data.table(panel="data", one.dt))+
  geom_point(aes(
    data.i+0.5, loss),
    shape=1,
    data=data.table(panel="loss", loss.dt))+
  facet_grid(panel ~ ., scales="free")
loss1 <- function(end){
  seg.dt <- loss.dt[end]
  gg+
    geom_vline(aes(
      xintercept=data.i+0.5),
      color=model.color,
      size=1,
      data=seg.dt)+
    geom_segment(aes(
      0.5, mean,
      xend=data.i+0.5, yend=mean),
      data=data.table(panel="data", seg.dt),
      size=1,
      color=model.color)
}
loss1(1)
```

--- 

# First step of dynamic programming

```{r}
loss1(10)
```

--- 

# First step of dynamic programming

```{r}
loss1(100)
```

--- 

# First step of dynamic programming

```{r}
loss1(nrow(one.dt))
```
