---
title: "Hidden Markov Models"
author: "Toby Dylan Hocking"
output: beamer_presentation
---

```{r opts, echo=FALSE}
knitr::opts_chunk$set(
  echo=FALSE, results=FALSE,
  fig.width=10,
  fig.height=6)
```

# Background: detecting abrupt changes is important 

Example from cancer diagnosis: breakpoints are associated with
aggressive disease in neuroblastoma.

```{r}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
})
data(neuroblastoma, package="neuroblastoma")
nb.dt <- data.table(neuroblastoma[["profiles"]])
setkey(nb.dt, profile.id, chromosome)
one.dt <- nb.dt[J("4", "2")]
ggplot()+
  scale_x_continuous(
    "Position/index in data sequence")+
  scale_y_continuous(
    "logratio (approximate DNA copy number)")+
  geom_point(aes(
    position, logratio),
    data=one.dt)
```

---

# Motivation for Hidden Markov Models (HMMs)

- Sometimes we have an interpretation / expectation of what the
  segments/clusters mean.
- For example in DNA copy number data the logratio=0 means normal copy
  number (two copies -- one from each parent), whereas higher logratio
  values indicate gain/amplification and lower values indicate
  loss/deletion.
- HMMs have a hidden state variable which is associated with each
  observed data variable in the sequence.
- There are typically a finite number of hidden states/clusters,
  $k\in\{1,\dots,K\}$.

---

# Baum-Welch learning algorithm

Let $K$ be the number of hidden states. For every state we have
- Let $A\in\mathbb [0,1]^{K\times K}$ be the transition matrix. Each
  entry $a_{ij}$ is the probability of of being in state $i$ at some
  time point $t$ and then at state $j$ at the next time point $t+1$.
- Let $\pi_k$ be the initial state distribution (prior weights). 
- There are also some emission distribution-specific parameters (mean,
  covariance) that are used to define the probablity of observing data
  $y$ in state $k$, $b_j(y)\in[0,1]$.
- These parameters are unknown in advance and must be learned from the
  data.
- Comparison with Gaussian Mixture Models: HMM has all of GMM
  parameters, plus transition matrix.

---

# Baum-Welch learning algorithm

Let $K$ be the number of hidden states. For every state we have
- Let $A\in\mathbb [0,1]^{K\times K}$ be the transition matrix. Each
  entry $a_{ij}$ is the probability of of being in state $i$ at some
  time point $t$ and then at state $j$ at the next time point $t+1$.
- Let $\pi_k$ be the initial state distribution (prior weights). 
- There are also some distribution-specific parameters (mean,
  covariance) that are used to define the probablity of observing data
  $y$ in state $k$, $b_j(y)\in[0,1]$.

---

# Computations

```{r}
n.states <- 3
set.seed(1)
rand.mat <- matrix(runif(n.states*n.states), n.states, n.states)
(A.mat <- rand.mat/rowSums(rand.mat))
mean.vec <- c(-1, 0, 1)*0.5
sd.param <- 1
rand.vec <- runif(n.states)
pi.vec <- rand.vec/sum(rand.vec)
y.vec <- one.dt[["logratio"]]
N.data <- length(y.vec)
alpha.mat <- matrix(NA, N.data, n.states)
alpha.mat[1,] <- pi.vec * dnorm(y.vec[1], mean.vec, sd.param)
for(data.t in 2:N.data){
  alpha.mat[data.t,] <- dnorm(y.vec[data.t], mean.vec, sd.param)*
    colSums(A.mat*matrix(alpha.mat[data.t-1,], n.states, n.states))
}
##http://bozeman.genome.washington.edu/compbio/mbt599_2006/hmm_scaling_revised.pdf
elnsum <- function(elnx, elny){
  fcase(
    elnx == -Inf, elny,
    elny == -Inf, elnx,
    elny < elnx, elnx + log(1+exp(elny-elnx)),
    elnx < elny, elny + log(1+exp(elnx-elny)))
}
elnproduct <- function(elnx, elny){
  ifelse(
    elnx == -Inf | elny == -Inf,
    -Inf,
    elnx+elny)
}
log.alpha.mat <- matrix(NA, N.data, n.states)
log.alpha.mat[1,] <- elnproduct(
  log(pi.vec), 
  dnorm(y.vec[1], mean.vec, sd.param, log=TRUE))
for(data.t in 2:N.data){
  log.alpha.vec <- rep(-Inf, n.states)
  for(state.i in 1:n.states){
    prod.vec <- elnproduct(
      log.alpha.mat[data.t-1,state.i],
      log(A.mat[state.i,]))
    log.alpha.vec <- elnsum(log.alpha.vec, prod.vec)
  }
  emission.prob.vec <- dnorm(y.vec[data.t], mean.vec, sd.param, log=TRUE)
  log.alpha.mat[data.t,] <- elnproduct(log.alpha.vec, emission.prob.vec)
}
diff.alpha.mat <- log.alpha.mat-log(alpha.mat)
head(diff.alpha.mat)
max(abs(diff.alpha.mat))

log.beta.mat <- matrix(NA, N.data, n.states)
log.beta.mat[N.data, ] <- 0
for(data.t in seq(N.data-1, 1)){
  for(state.j in 1:n.states){
  }
}

```
