---
title: "Principal Components Analysis"
author: "Toby Dylan Hocking"
output: beamer_presentation
---

```{r opts, echo=FALSE}
knitr::opts_chunk$set(
  echo=FALSE, results=FALSE,
  fig.width=10,
  fig.height=7)
```

# Background/motivation: dimensionality reduction

- High dimensional data are difficult to visualize.
- For example each observation/example in the zip data is of dimension
  16 x 16 = 256 pixels.
- We would like to map each observation into a lower-dimensional
  space for visualization / understanding patterns in the data.
  
# Example 1: 2d iris data

```{r}
## 0. iris data scatterplot two columns with one principal component line
##    and orthogonal projection segments.
library(data.table)
i.df <- iris[, 2:3]
str(i.df)
library(ggplot2)
gg <- ggplot()+
  geom_point(aes(
    x=Petal.Length, y=Sepal.Width),
    shape=1,
    data=i.df)+
  coord_equal()
gg
```

---

# Project 2d data onto 1d subspace (line)

Why this line? 

```{r}
pc.fit <- prcomp(i.df)
str(pc.fit)
pc.fit[["rotation"]]
lambda.vec <- seq(-3, 3, by=0.5)
##lambda.vec <- pc.fit$x[,1] ## this is lambda from the book!
line.dt.list <- list()
PC1 <- pc.fit[["rotation"]][,1]
PC2 <- pc.fit[["rotation"]][,2]
f <- function(l)mean.vec + l * PC1
mean.vec <- colMeans(i.df)
for(lambda in lambda.vec){
  line.vec <- f(lambda)
  line.dt.list[[paste(lambda)]] <- data.table(lambda, t(line.vec))
}
line.dt <- do.call(rbind, line.dt.list)
line.dt[, method := "grid"]
##y= mx + b
## Sepal.Width = slope * Petal.Length + intercept
## What is the intercept? that is b where 
## f(Petal.Length) = 0 = mean(Petal.Length) + lambda * PC1(Petal.Length)
## f(Sepal.Width) = b = mean(Sepal.Width)  + lambda * PC1(Sepal.Width)
## lambda = -mean(Sepal.Width) / PC1(Sepal.Width)
lambda.intercept <- -mean.vec[["Petal.Length"]]/PC1[["Petal.Length"]]
f(lambda.intercept)
abline.dt <- data.table(
  method="slope/intercept",
  ## f(Petal.Length) =
  ## mean(Petal.Length)
  ## - mean(Sepal.Width)*PC1(Petal.Length)/PC1(Sepal.Width)
  ## + f(Sepal.Width)*PC1(Petal.Length)/PC1(Sepal.Width)
  intercept=mean.vec[["Sepal.Width"]] +
    lambda.intercept*PC1[["Sepal.Width"]],
  slope=PC1[["Sepal.Width"]]/PC1[["Petal.Length"]])
## [f(Sepal.Width) - mean(Sepal.Width)]/PC1(Sepal.Width) = lambda
## f(Petal.Length) = mean(Petal.Length) +
## [(f(Sepal.Width)-mean(Sepal.Width))/PC1(Sepal.Width)] * PC1(Petal.Length)
gg+
  geom_abline(aes(
    slope=slope,
    intercept=intercept),
    data=abline.dt)
```

---

# Principal Components Projection

The first principal component is the line which minimizes the
reconstruction error, squared distance between projection and data.

```{r}
PC1.mat <- matrix(PC1, nrow=nrow(iris), ncol=2, byrow=TRUE)
mean.vec <- colMeans(i.df)
mean.mat <- matrix(mean.vec, nrow=nrow(iris), ncol=2, byrow=TRUE)
pred.mat <- mean.mat + PC1.mat * pc.fit[["x"]][, 1] #lambda from book.
colnames(pred.mat) <- colnames(i.df)
both.dt <- rbind(
  data.table(i.df, type="data", Species=iris$Species),
  data.table(pred.mat, type="projection",Species=iris$Species))
segs.dt <- data.table(data=i.df, pred=pred.mat)
ggplot()+
  coord_equal()+
  geom_point(aes(
    x=Petal.Length, y=Sepal.Width, fill=type),
    shape=21,
    size=2,
    data=both.dt)+
  geom_abline(aes(
    slope=slope,
    intercept=intercept),
    data=abline.dt)+
  geom_blank(aes(
    Petal.Length, Sepal.Width),
    data=line.dt)+
  geom_segment(aes(
    y=data.Sepal.Width,
    yend=pred.Sepal.Width,
    x=data.Petal.Length,
    xend=pred.Petal.Length),
    data=segs.dt)
```

---

# Map label onto projection

```{r}
color.code <- c(
  setosa="#1B9E77",
  versicolor="#D95F02",
  virginica="#7570B3",
  "1"="#E7298A",
  "2"="#66A61E",
  "3"="#E6AB02", 
  "4"="#A6761D")
ggplot()+
  coord_equal()+
  scale_fill_manual(values=color.code)+
  geom_segment(aes(
    y=data.Sepal.Width,
    yend=pred.Sepal.Width,
    x=data.Petal.Length,
    xend=pred.Petal.Length),
    data=segs.dt)+
  geom_blank(aes(
    Petal.Length, Sepal.Width),
    data=line.dt)+
  geom_point(aes(
    x=Petal.Length, y=Sepal.Width, fill=Species),
    shape=21,
    size=2,
    data=both.dt)+
  geom_abline(aes(
    slope=slope,
    intercept=intercept),
    data=abline.dt)
```

---

# Map label onto projection

```{r}
uniq.data <- unique(both.dt[type=="data", .(Petal.Length, Sepal.Width)])
ggplot()+
  coord_equal()+
  scale_fill_manual(values=color.code)+
  geom_point(aes(
    x=Petal.Length, y=Sepal.Width, fill=Species),
    shape=21,
    size=2,
    data=both.dt[type=="projection"])+
  geom_point(aes(
    x=Petal.Length, y=Sepal.Width),
    alpha=0.3,
    data=uniq.data)+
  geom_blank(aes(
    Petal.Length, Sepal.Width),
    data=line.dt)+
  geom_abline(aes(
    slope=slope,
    intercept=intercept),
    data=abline.dt)
```

---

# Principal component 1, amount along projection

```{r, fig.height=5}
ggplot()+
  coord_equal()+
  scale_fill_manual(values=color.code)+
  ggforce::geom_circle(aes(
    x0=Petal.Length, y0=Sepal.Width, r=1),
    fill="grey50",
    alpha=0.5,
    data=data.table(t(mean.vec)))+
  geom_point(aes(
    x=Petal.Length, y=Sepal.Width),
    alpha=0.3,
    data=uniq.data)+
  geom_point(aes(
    x=Petal.Length, y=Sepal.Width, fill=Species),
    shape=21,
    size=2,
    data=both.dt[type=="projection"])+
  geom_blank(aes(
    x=Petal.Length, y=Sepal.Width),
    data=both.dt[type=="data"])+
  geom_abline(aes(
    slope=slope,
    intercept=intercept),
    data=abline.dt)+
  geom_segment(aes(
    mean.Petal.Length, mean.Sepal.Width,
    xend=dir.Petal.Length, yend=dir.Sepal.Width),
    size=2,
    arrow=grid::arrow(length=grid::unit(0.2, "in"), type="closed"),
    data=data.table(t(c(mean=mean.vec, dir=mean.vec+PC1))))+
  geom_label(aes(
    Petal.Length, Sepal.Width, label=lambda),
    alpha=0.7,
    data=line.dt)
```

- 0 represents mean of data.
- $0\rightarrow 1$ represents an orthogonal unit vector.

---

# Principal component 2

```{r, fig.height=5}
ggplot()+
  coord_equal()+
  scale_fill_manual(values=color.code)+
  ggforce::geom_circle(aes(
    x0=Petal.Length, y0=Sepal.Width, r=1),
    fill="grey50",
    alpha=0.5,
    data=data.table(t(mean.vec)))+
  geom_point(aes(
    x=Petal.Length, y=Sepal.Width),
    alpha=0.3,
    data=uniq.data)+
  geom_blank(aes(
    x=Petal.Length, y=Sepal.Width),
    data=both.dt[type=="data"])+
  geom_abline(aes(
    slope=slope,
    intercept=intercept),
    data=abline.dt)+
  geom_segment(aes(
    mean.Petal.Length, mean.Sepal.Width,
    xend=dir.Petal.Length, yend=dir.Sepal.Width),
    size=2,
    arrow=grid::arrow(length=grid::unit(0.2, "in"), type="closed"),
    data=data.table(t(c(mean=mean.vec, dir=mean.vec+PC2))))
```

---

# Re-plot using PC units

```{r}
ggplot()+
  scale_fill_manual(values=color.code)+
  geom_point(aes(
    PC1, PC2, fill=Species),
    shape=21,
    data=data.table(pc.fit$x, Species=iris$Species))+
  coord_equal()
```

---

# Mathematical representation

Each of the $n$ inputs $x_i \in \mathbb R^p$ where $p$ is the input
dimension, $p=2$ for iris in previous slides, $p=784$ for images in
next slides.

$$ \min \sum_{i=1}^n || x_i - \mu - V_q \lambda_i ||^2. $$

- $\mu\in\mathbb R^p$ is mean vector.
- $V_q\in\mathbb R^{p\times q}$ is an orthogonal matrix (each column
  is an orthogonal unit vector).
- $\lambda_i\in\mathbb R^q$ is a vector of principal components
  (contribution of each unit vector).

---

# Same analysis with MNIST digit data

```{r}
## 1. MNIST scatterplot of digit labels on first two pca directions (all
## classes).
if(file.exists("figure-fashion-mnist-data.rds")){
  data.list <- readRDS("figure-fashion-mnist-data.rds")
}else{
  data.list <- list(
    fashion=keras::dataset_fashion_mnist(),
    digits=keras::dataset_mnist())
  saveRDS(data.list, "figure-fashion-mnist-data.rds")
}
str(data.list$digits)

X.mat.list <- list()
obs.i.vec <- 1:1000
for(obs.i in obs.i.vec){
  X.mat.list[[obs.i]] <- as.numeric(data.list$digits$train$x[obs.i, , ])
}
X.mat <- do.call(rbind, X.mat.list)
digit <- data.list$digits$train$y[obs.i.vec]
## specifying rank arg is faster, use it if you don't need all the
## principal components.
digits.fit <- prcomp(X.mat, rank=2)
digits.pc.dt <- data.table(
  digits.fit$x, digit=factor(digit))

ggplot()+
  geom_text(aes(
    PC1, PC2, label=digit),
    data=digits.pc.dt)
```

---

# Alternate visualization

```{r}
ggplot()+
  geom_point(aes(
    PC1, PC2, fill=digit),
    shape=21,
    data=digits.pc.dt)+
  scale_fill_brewer(palette="Set3")
```

---

# Another PCA on just one digit class

```{r}
## specifying rank arg is faster, use it if you don't need all the
## principal components.
is.one.digit <- digit==3
X.one.digit <- X.mat[is.one.digit,]
y.one.digit <- digit[is.one.digit]
table(y.one.digit)
digits.fit <- prcomp(X.one.digit)

tile.dt.list <- list()
n.pixels <- 28
for(digit.i in 1:nrow(X.one.digit)){
  tile.dt.list[[digit.i]] <- data.table(
    digit.i,
    intensity=X.one.digit[digit.i,],
    row=rep(1:n.pixels, n.pixels),
    col=rep(1:n.pixels, each=n.pixels))
}
tile.dt <- do.call(rbind, tile.dt.list)
ggplot()+
  scale_fill_gradient(low="black", high="white")+
  geom_tile(aes(
    col, -row, fill=intensity),
    data=tile.dt)+
  coord_equal()+
  facet_wrap("digit.i", ncol=15)
```

---

# Mapping onto first two PCs

```{r}
digits.pc.mat <- digits.fit[["x"]][, 1:2]
digits.pc.dt <- data.table(digits.pc.mat)
digits.pc.dt[, digit.i := 1:.N]
line.dt.list <- list()
expand.grid.args <- list()
for(pc.name in colnames(digits.pc.mat)){
  pc.vec <- digits.pc.dt[[pc.name]]
  n.grid <- 6
  pc.grid <- seq(min(pc.vec), max(pc.vec), l=n.grid)[-c(1,n.grid)]
  expand.grid.args[[pc.name]] <- pc.grid
  line.dt.list[[pc.name]] <- data.table(
    pc.name,
    type="grid",
    pc.grid)
}
grid.point.dt <- data.table(do.call(expand.grid, expand.grid.args))
grid.point.dt[, type := "grid"]
digits.pc.dt[, type := "data"]
gg.grid <- ggplot()+
  geom_point(aes(
    PC1, PC2, color=type),
    data=digits.pc.dt)+
  geom_point(aes(
    PC1, PC2, color=type),
    data=grid.point.dt)+
  geom_vline(aes(
    xintercept=pc.grid, color=type),
    data=line.dt.list[["PC1"]])+
  geom_hline(aes(
    yintercept=pc.grid, color=type),
    data=line.dt.list[["PC2"]])
gg.grid
```

---

# Highlight closest data point to each grid point

```{r}
grid.tile.dt.list <- list()
closest.pc.list <- list()
for(grid.i in 1:nrow(grid.point.dt)){
  one.grid.point <- grid.point.dt[grid.i]
  one.grid.mat <- one.grid.point[, matrix(
    c(PC1, PC2), nrow=nrow(digits.pc.mat), ncol=2, byrow=TRUE)]
  dist.vec <- rowSums((one.grid.mat-digits.pc.mat)^2)
  closest.i <- which.min(dist.vec)
  closest.pc.list[[grid.i]] <- data.table(digits.pc.mat[closest.i,,drop=FALSE])
  grid.tile.dt.list[[grid.i]] <- data.table(
    one.grid.point,
    grid.i,
    intensity=X.one.digit[grid.i,],
    row=rep(1:n.pixels, n.pixels),
    col=rep(1:n.pixels, each=n.pixels))
}
grid.tile.dt <- do.call(rbind, grid.tile.dt.list)
closest.pc <- do.call(rbind, closest.pc.list)
gg.grid+
  geom_point(aes(
    PC1, PC2),
    data=closest.pc)
```

---

# Digits highlighted

```{r}
ggplot()+
  geom_tile(aes(
    col, -row, fill=intensity),
    data=grid.tile.dt)+
  scale_fill_gradient(low="black", high="white")+
  coord_equal()+
  facet_grid(PC2 ~ PC1, labeller=label_both)
```

---

# One digit 

```{r}
## 2. mean and first two eigendigits. reconstruction % and number of
##    components in different panels for a single example. (MLAPP Fig
##    12.6)
obs.i <- 11
obs.i.tiles <- data.table(
  intensity=X.one.digit[obs.i,],
  row=rep(1:n.pixels, n.pixels),
  col=rep(1:n.pixels, each=n.pixels))
n.pixels <- 28
reconstruction.dt.list <- list()
for(n.components in 0:nrow(digits.fit$x)){
  reconstruction <- if(n.components==0){
    digits.fit$center
  }else{
    reconstruction +
      digits.fit$rotation[,n.components] *
      digits.fit$x[obs.i,n.components]
  }
  reconstruction.dt.list[[paste(n.components)]] <- data.table(
    intensity=reconstruction,
    row=rep(1:n.pixels, n.pixels),
    col=rep(1:n.pixels, each=n.pixels))
}
reconstruction.dt <- do.call(rbind, reconstruction.dt.list)
rec.range <- range(reconstruction.dt$intensity)
plot.components <- function(comp){
  ggplot()+
    coord_equal()+
    geom_tile(aes(
      col, -row, fill=intensity),
      data=reconstruction.dt.list[[paste(comp)]])+
    scale_fill_gradient(low="black", high="white", limits=rec.range)
}
ggplot()+
  coord_equal()+
  geom_tile(aes(
    col, -row, fill=intensity),
    data=obs.i.tiles)+
  scale_fill_gradient(low="black", high="white", limits=rec.range)
```

---

# Reconstruction with no components (mean)

```{r}
plot.components(0)
```

---

# Reconstruction with one PC

```{r}
plot.components(1)
```

---

# Reconstruction with 2 PCs

```{r}
plot.components(2)
```

---

# Reconstruction with 5 PCs

```{r}
plot.components(5)
```

---

# Reconstruction with 10 PCs

```{r}
plot.components(10)
```

---

# Reconstruction with 50 PCs

```{r}
plot.components(50)
```

---

# Reconstruction with all PCs

```{r}
plot.components(rev(names(reconstruction.dt.list))[1])
```

---

# How to compute PCA?

SVD = Singular Value Decomposition (many algorithms available to
compute).

$$ X = UDV^T $$

- $X \in \mathbb R^{n\times p}$ data matrix.
- $U \in \mathbb R^{n\times p}$ orthogonal matrix.
- $D \in \mathbb R^{p\times p}$ diagonal matrix.
- $V \in \mathbb R^{p\times p}$ orthogonal matrix.
- The $V_q$ we want for PCA is the first $q$ columns of $V$.
- The columns of $UD$ are the principal components, $\lambda_i$
  values.

---

# Possible exam questions

- TODO
